A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-28 podc_paper/algorithm.tex
working on code

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{?}.
We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
We use a similar abbreviation \var{num\sub{deq}}.

@@

\begin{figure}
\begin{minipage}[t]{0.46\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \var{Block} with \var{element} \assign\ \var{null},\\
	    \var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}},\\
	    \var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}+1}}
    \State \Call{Append}{n}
    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf.\head, 1}
    \State \Return{ \Call{FindResponse}{b, i}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\var{Node v}}
    \State \linecomment{propagates blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{v} \label{secondRefresh}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{v.parent}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\var{Node v}}
    \State \linecomment{tries to append a new block to \var{v.blocks}}
    \State \var{h \assign\ v.\head} \label{readHead}
    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}} \label{startHelpChild1}
        \State \var{h\sub{dir} \assign\ v.dir.\head} \label{readChildHead}
        \If{\var{v.dir.blocks[h\sub{dir}] $\neq$ null}} \label{ifHeadnotNull}
            \State{\Call{Advance}{v.dir, h\sub{dir}}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
    \If{\var{new.num = 0}} \Return{\var{true}} \label{addOP} 
    \EndIf
    \State{\var{result} \assign\ \var{CAS(v.blocks[h], null, new)}} \label{cas}
    \State{\var{\Call{Advance}{v, h}}} \label{advance}
    \State \Return{ \tt{result}}
\EndFunction{Refresh}

@@

\Function{void}{Advance}{\var{Node v}, \var{int h}} 
    \State \linecomment{set \nf{v.blocks[h].super} and }
    \State \linecomment{increment \nf{v.head} from \nf{h} to \nf{h+1}}
    \State \var{h\sub{p}} \assign\ \var{v.parent.\head} \label{readParentHead}
    \State \var{CAS(v.blocks[h].super, null, h\sub{p})} \label{setSuper1}
    \State \var{CAS(v.head, h, h+1)} \label{incrementHead}
\EndFunction{Advance}

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.53\textwidth}

@@

\Function{Block}{CreateBlock}{\var{Node v, int i}} 
    \State\linecomment{create new block to be installed in \var{v.blocks[i]}}
    \State let \var{new} be a new Block \label{initNewBlock}
    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}}
        \State \var{index\sub{prev} \assign\ v.blocks[i-1].\edir} \label{prevLine}
        \State \var{new.\edir\ \assign\ v.dir.\head-1} \label{lastLine}
        \State \linecomment{\var{new} contains \var{v.dir.blocks[index\sub{prev}+1..new.\edir]}}
        \State \var{block\sub{prev} \assign\ v.dir.blocks[index\sub{prev}]}
        \State \var{block\sub{last} \assign\ v.dir.blocks[new.\edir]}
    \EndFor
	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{enq} + \\
			v.right.blocks[\eright].sum\sub{enq}}}
	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[\eleft].sum\sub{deq} + \\
			v.right.blocks[\eright].sum\sub{deq}}}
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new.size \assign\ max(0, \\ v.blocks[i-1].\size\ + new.num\sub{enq} - new.num\sub{deq})}\label{computeLength}}
    \EndIf
    \State \Return \tt{new}
\EndFunction{CreateBlock}

@@

\Function{element}{FindResponse}{\var{int b, int i}}
\State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
\If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
\State \linecomment{queue is empty when the dequeue is linearized}
\State \Return \var{null} \label{returnNull}
\Else \ \linecomment{response is the \var{e}th enqueue in the root}
\State \var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - root.blocks[b-1].size}\label{computeE}
\State \Return \var{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}
\caption{\tt{\sl{Queue}}}
\end{figure}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b}) \assign\ D_\tt{y}(\tt{root},\tt{x})$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir\assign\ (parent.left=n ? left: right)} 
\State \tt{superblockIndex \assign\ parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
\Else \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
\EndIf
\EndFunction{IndexDequeue}

>>>>

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode}.
We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
We use \var{num\sub{deq}} similarly.
\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\begin{figure}
\begin{minipage}[t]{0.465\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{n} be a new \var{Block} with \var{element} \assign\ \var{null},\\
	    \var{sum\sub{enq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{enq}},\\
	    \var{sum\sub{deq} \assign\ leaf.blocks[leaf.\head-1].sum\sub{deq}+1}}
    \State \Call{Append}{n}
    \State \var{<b, i>} \assign\ \Call{IndexDequeue}{leaf, leaf.\head, 1}
    \State \Return{ \Call{FindResponse}{b, i}}
    \label{deqRest}
\EndFunction{Dequeue}

@@

\Function{void}{Propagate}{\var{Node v}}
    \State \linecomment{propagate blocks from \var{v}'s children to root}
    \If{\bf{not} \Call{Refresh}{v}} \label{firstRefresh}  \hfill \com\ double refresh
        \State \Call{Refresh}{v} \label{secondRefresh}
    \EndIf
    \If{\var{v} \bf{is not} \var{root}} \hfill \com\ recurse up tree
        \State \Call{Propagate}{v.parent}
    \EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\var{Node v}}
    \State \linecomment{try to append a new block to \var{v.blocks}}
    \State \var{h \assign\ v.\head} \label{readHead}
    \ForEach{\var{dir} {\keywordfont{in}} \var{\{left, right\}}} \label{startHelpChild1}
        \State \var{h\sub{dir} \assign\ v.dir.\head} \label{readChildHead}
        \If{\var{v.dir.blocks[h\sub{dir}] $\neq$ null}} \label{ifHeadnotNull}
            \State{\Call{Advance}{v.dir, h\sub{dir}}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{v,h} \label{invokeCreateBlock}
    \If{\var{new = null}} \Return{\var{true}} \label{addOP} 
    \EndIf
    \State{\var{result} \assign\ \var{CAS(v.blocks[h], null, new)}} \label{cas}
    \State{\var{\Call{Advance}{v, h}}} \label{advance}
    \State \Return{ \tt{result}}
\EndFunction{Refresh}

@@

\Function{void}{Advance}{\var{Node v, int h}} 
    \State \linecomment{set \nf{v.blocks[h].super} and }
    \State \linecomment{increment \nf{v.head} from \nf{h} to \nf{h+1}}
    \State \var{h\sub{p}} \assign\ \var{v.parent.\head} \label{readParentHead}
    \State \var{CAS(v.blocks[h].super, null, h\sub{p})} \label{setSuper1}
    \State \var{CAS(v.head, h, h+1)} \label{incrementHead}
\EndFunction{Advance}

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.525\textwidth}

@@

\Function{Block}{CreateBlock}{\var{Node v, int i}} 
    \State\linecomment{create new block for a \op{Refresh} to install in \var{v.blocks[i]}}
    \State\linecomment{(return \var{null} if no blocks need to be propagated to \var{v})}
    \State let \var{new} be a new \var{Block} \label{initNewBlock}
    \State \var{new.\eleft} \assign\ $\var{v.left.head} - 1$
    \State \var{new.\eright} \assign\ $\var{v.right.head} - 1$
	\State \hangbox{\var{new.sum\sub{enq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{enq} + \\
			v.right.blocks[new.\eright].sum\sub{enq}}}
	\State \hangbox{\var{new.sum\sub{deq}} \assign\ \var{v.left.blocks[new.\eleft].sum\sub{deq} + \\
			v.right.blocks[new.\eright].sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ \var{new.sum\sub{enq} - v.blocks[i-1].sum\sub{enq}}
    \State \var{num\sub{deq}} \assign\ \var{new.sum\sub{deq} - v.blocks[i-1].sum\sub{deq}}
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new.size \assign\ max(0, v.blocks[i-1].\size\ + num\sub{enq} - num\sub{deq})}\label{computeLength}}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}
        \State \Return \var{null}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{<int, int>}{IndexDequeue}{\var{Node v, int b, int i}}
    \State \linecomment{return \var{<x, y>} such that \var{i}th dequeue in D(\var{v.blocks[b]})}
    \State \linecomment{is \var{y}th dequeue of D(\var{root.blocks[x]})}
    \State \linecomment{Precondition: \var{v.blocks[b]} has been propagated to root}
    \State \linecomment{and $\var{v.blocks[b].num\sub{deq}}\geq \var{i}$}
    \If{$\var{v} = \var{root}$}
        \State\Return \var{<b, i>} \label{indexBaseCase}
    \Else
	    \State \var{dir \assign\ (v.parent.left = v ? left : right)} 
    	\State \var{sup \assign\ v.blocks[b].super}
	    \If{$\var{b} > \var{v.parent.blocks[sup].end\sub{dir}}$}
	        \State{\var{sup} \assign\ $\var{sup}+1$}
	    \EndIf
	    \State \hangbox{\var{i += v.blocks[b-1].sum\sub{deq} - \\
	    		v.blocks[v.parent.blocks[sup-1].end\sub{dir}].sum\sub{deq}}}
        \If{$\var{dir} = \var{right}$} 
        	\State \hangbox{\var{i += v.blocks[v.parent.blocks[sup].end\sub{left}].sum\sub{deq} -\\
					v.blocks[v.parent.blocks[sup-1].end\sub{left}].sum\sub{deq}}\label{considerLeftBeforeRight}}
        \EndIf \label{computeISuperEnd}
        \State \Return\Call{IndexDequeue}{v.parent, sup, i}
    \EndIf
\EndFunction{IndexDequeue}

\Function{element}{FindResponse}{\var{int b, int i}}
    \State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
    \If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
        \State \linecomment{queue is empty when the dequeue is linearized}
        \State \Return \var{null} \label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \hangbox{\var{e} \assign\ \var{i + root.blocks[b-1].sum\sub{enq} - \\
	root.blocks[b-1].size}\label{computeE}}
        \State \Return \var{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}
\caption{Queue implementation.\label{pseudocode}}
\end{figure}

@@
